Physcraper: A Python package for continually updated phylogenetic trees using the Open Tree of Life

### 2021-04-02 sources/bmc_template_submission/bmc_article.tex
some edits

%%%%%%%%%%%%%%%%
%% Background %%
%%
\section*{Background}
Phylogenies capture the shared history of organisms and provide key evolutionary
context for our biological observations \cite{dobzhansky1973nothing}.
Updating existing phylogenies with publicly available molecular sequence data provides
the opportunity to study
the evolutionary history of many taxa in a reproducible and continuous manner.
Here, we introduce Physcraper, a tool that establishes a data interoperability
framework for biological databases to automate data connections across databases.
Physcraper's main goal is to build upon published alignments to extend existing
phylogenetic inferences with more data and taxa, which have been proved to improve
phylogenetic reconstruction \cite{hillis1996inferring, natsidis2019phylogenomics},
time of divergence estimates \cite{schulte2013undersampling, soares2015influence} and
biogeographic analyses \cite{kayaalp2017back},
as well as help in resolving phylogenetic conflict \cite{hedtke2006resolution, townsend2010optimal, natsidis2019phylogenomics}.
Physcraper updates a starting tree and single locus alignments
with public DNA data, and links the tips in these trees to a unified, interoperable
taxonomic resource \cite{rees2017automated}.

@@

Decades of single locus sequencing have generated
massive amounts of homologous DNA datasets that have the potential to be used for phylogenetic
reconstruction at many scales \cite{sanderson2008phylota}.
More than a decade ago, GenBank release 159 (April 15, 2007) already hosted 72
million DNA sequences that were gauged to have the potential to resolve phylogenetic
relationships of 98.05\% of the almost 241,000
distinct taxa in the NCBI taxonomy at the time \cite{sanderson2008phylota}.
However, even thirteen years later, phylogenetic estimates for many of these taxa are not available \cite{mctavish_how_2017}.
OpenTree's comprehensive tree of life comprises 2.3 million tips, of which around
90,000 are supported by phylogenies - the remaining 1.4 million taxa are placed
in the tree based on taxonomy.
There is a considerable amount of phylogenetically
informative data in GenBank with the potential to fill these phylogenetic gaps in
the tree of life, but this data either has not been analysed or the analyses
have not been made publicly available \cite{mctavish_how_2017}.

@@

Physcraper improves on previous work in automating phylogenetic reconstruction by leveraging
the power of existing homology hypotheses that taxon specialists have assessed
and deemed appropriate for a specific phylogenetic scope.
There are almost 8,200 publicly available, peer-reviewed curated alignments, covering
around 100,000 distinct taxa in the TreeBASE database \cite{piel2009treebase}, which
can be leveraged as seeds to mine molecular databases, and as ``jump-start'' alignments
for phylogenetic reconstructions \cite{morrison2006multiple} to continually enrich,
update and compare existing phylogenetic knowledge.

@@

Physcraper is implemented with Python and can also be run on the command line.
It currently consists of 12 modules.
%% , classes, methods and functions.%%
For testing and improving the quality of the Python code syntax in the Phsycraper code,
we used the Pylint \cite{pylintWeb, pylintManual} software with a configuration file.
We improved the code syntax on modules with low Pylint scores and fixed errors by
following Pylint's recommendations.
In some cases, we overruled Pylint's recommendations by using its check-disabling
system. As of now, all modules have a Pylint score of 10/10.

>>>>

%%%%%%%%%%%%%%%%
%% Background %%
%%
\section*{Background}
Phylogenies capture the shared history of organisms and provide key evolutionary
context for our biological observations \cite{dobzhansky1973nothing}.
Updating existing phylogenies with publicly available molecular sequence data provides
the opportunity to study
the evolutionary history of many taxa in a reproducible and continuous manner.
Here, we introduce Physcraper, a tool that establishes a data interoperability
framework for biological databases to automate data connections across databases.
Physcraper's main goal is to build upon published alignments to extend existing
phylogenetic inferences with more data and taxa, which improve
phylogenetic reconstruction \cite{hillis1996inferring, natsidis2019phylogenomics},
time of divergence estimates \cite{schulte2013undersampling, soares2015influence},
biogeographic analyses \cite{kayaalp2017back},
as well as help in resolving phylogenetic conflict \cite{hedtke2006resolution, townsend2010optimal, natsidis2019phylogenomics}.
Physcraper updates a starting single locus alignment and corresponding tree
with public DNA data, and links the tips in the updated trees to a unified, interoperable
taxonomic resource \cite{rees2017automated}.

@@

Decades of single locus sequencing have generated
massive amounts of homologous DNA datasets that have the potential to be used for phylogenetic
reconstruction at many scales \cite{sanderson2008phylota}.
More than a decade ago, GenBank release 159 (April 15, 2007) already hosted 72
million DNA sequences that were gauged to have the potential to resolve phylogenetic
relationships of 98.05\% of the almost 241,000
distinct taxa in the NCBI taxonomy at the time \cite{sanderson2008phylota}.
However, even thirteen years later, phylogenetic estimates for most of these taxa
are still not available \cite{mctavish_how_2017}.
The Open Tree of Life project assembles a comprehensive synthetic tree of life comprising
2.3 million tips, of which around only
90,000 are supported by phylogenetic data - the remaining 1.4 million taxa are placed
in the tree based on taxonomy.
There is a considerable amount of phylogenetically
informative data in GenBank with the potential to fill these phylogenetic gaps in
the tree of life, but this data either has not been analysed or the analyses
have not been made publicly available \cite{mctavish_how_2017}.

@@

Physcraper improves on previous work in automating phylogenetic reconstruction by leveraging
the power of existing homology hypotheses that taxon specialists have assessed
and deemed appropriate for a specific phylogenetic scope.
There are almost 8,200 publicly available, peer-reviewed curated alignments, covering
around 100,000 distinct taxa in the TreeBASE database \cite{piel2009treebase}, which
can be leveraged as seeds to mine molecular databases, and as "jump-start" alignments
for phylogenetic reconstructions \cite{morrison2006multiple} to continually enrich,
update and compare existing phylogenetic knowledge.

@@

Physcraper is implemented with Python and can also be run on the command line.
It currently consists of 12 modules.
%% , classes, methods and functions.%%
For testing and improving the quality of the Python code syntax in the Phsycraper code,
we used the Pylint \cite{pylintWeb, pylintManual} software with a configuration file.
We improved the code syntax on modules with low Pylint scores and fixed errors by
following Pylint's recommendations.
Based on Physcraper's software design choices, some of Pylint's recommendations were overruled
by using its check-disabling system, and are explained in the code.
As of now, all modules have a Pylint score of 10/10.
